---
title: "Mode Hunting Using Beta-Trees"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mode Hunting Using Beta-Trees}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(igraph)
library(mvtnorm)
library(BetaTree)
```

In this vignette, we will demonstrate how to use a Beta-tree histogram to identify modes in a continuous probability distribution. 

The mode of a continuous probability distribution is a local maxima of the probability density. For example, the following plot shows the density contours of a mixture of 2-dim Gaussian  and the asterisks indicate the two modes of the density function. We are often interested to identify modes in a distribution because they indicate sub-populations (here, the two modes indicate the two mixture components). 

```{r, echo = F}
set.seed(1)
```

```{r, echo = F}
w <- c(0.4, 0.6)
mu12d <- c(-1.5, 0.6)
mu22d <- c(2, -1.5)
sigma12d <-  matrix(c(1, 0.5, 0.5, 1), 2, 2, byrow = T)
sigma22d <-  matrix(c(1, 0, 0, 1), 2, 2, byrow = T)

# the density function: dmvnornm can apply to rows of a matrix
densityfun <- function(x, y){
  t <- cbind(x, y)
  
  w[1] * dmvnorm(t, mean = mu12d, sigma = sigma12d, log = F) +
    w[2] * dmvnorm(t, mean = mu22d, sigma = sigma22d, log = F)
}

```

```{r, echo = F, fig.width = 4, fig.height = 3, fig.align='center'}
x <- seq(-6, 6, 0.1)
y <- seq(-6, 4, 0.1)

val <- expand.grid(x, y)
z <- densityfun(val[,1], val[,2])

ggplot() + 
  geom_contour(aes(x = val[,1], y = val[,2], z = z), color = "black", size = 0.8) + 
  geom_point(aes(x = c(-1.5, 2), y= c(0.6, -1.5)),  shape = 8, color = "red", size = 2) + 
  xlab("X") + 
  ylab("Y") + 
  theme_bw() 

```

For a histogram, we can think of a mode as a region whose average density is higher compared to its neighbors. Put in a different way, suppose two regions $R_1$ and $R_2$ are distinct modes, then along any path connecting $R_1$ and $R_2$ there exists at least one region $R$ whose average density is lower than those of both $R_1$ and $R_2$. 

## Example: Detecting modes in a mixture of 2-dim Gaussian

```{r, echo = F}
n <- 2000 # num of obs.
d <- 2 # num of dim.
```

In this example, we simulate $n=$ `r n` obs. from a 2-dim Gaussian mixture we showed before 

$$
\frac{2}{5} {N}\left(\left(\begin{matrix}-1.5 \\0.6\end{matrix}\right),\left(\begin{matrix}1 & 0.5 \\ 0.5 & 1\end{matrix}\right) \right) + \frac{3}{5} {N}\left(\left(\begin{matrix}2 \\-1.5\end{matrix}\right),\left(\begin{matrix}1 & 0 \\ 0 & 1\end{matrix}\right) \right), 
$$
```{r, echo = F}
mu2d <- list(c(-1.5, 0.6), c(2, -1.5)) # mean 
sigma12d <- matrix(c(1, 0.5, 0.5, 1), d, d, byrow = T)
sigma22d <- matrix(c(1, 0, 0, 1), d, d, byrow = T)
sigma2d <- list(sigma12d, sigma22d) # covariance

N <- apply(rmultinom(n, 1, prob = c(0.4, 0.6)), 2, function(t) which(t == 1)) 
nn <- table(N) # number of elements from each mixture component
X <- matrix(0, n, d) # observations
# sample from each of the mixture
for(i in 1:2){
  R <- chol(sigma2d[[i]])
  G <- matrix(rnorm(nn[i] * 2, 0, 1), nn[i], d) %*% R
  X[which(N == i), ] <-   t(t(G) + mu2d[[i]])
}
```


```{r, echo = F, fig.show='hide', fig.height = 3, fig.width= 4, fig.align='center'}
ggplot() + 
  geom_point(aes(x = X[,1], y = X[,2] , color = as.factor(N)), size = 0.5) + 
  theme_bw() + 
  labs(color = "Component", x = "X", y="Y") + 
  theme(legend.position = c(0.85, 0.8)) 
```

We will start by creating a Beta-tree histogram. Here, we set the confidence level `alpha = 0.1` and we use weighted bonferroni method for multiple testing correction. The densities are higher in the upper left and the lower right regions, suggesting there are two modes in the distribution. 

```{r, fig.height = 3, fig.width= 5, fig.align='center'}
hist <- BuildHist(X, alpha = 0.1, method = "weighted_bonferroni", plot= T)
```

We use the `FindModes()` function to identify modes in a Beta-trees histogram. This function has three parameters, the first is the Beta-tree histogram (output of the `BuildHist` function), the second is the data dimension `d`, and the third is the cutoff value of the path length (we only check paths whose length is at most `cutoff`). We set `cutoff=1000` here, which means we will look at every path between two regions. However, it might be infeasible to check every path because of computation restraints, so typically we would set a smaller cutoff, e.g., `cutoff = 6`. 

```{r}
modes <- FindModes(hist = hist, d = d, cutoff = 1000)
```

The `FindModes` function returns a list with three components: `mode` returns the index of the modes, `hist` returns the input histogram, and `g` returns the adjacency graph of regions in the histogram. 

```{r}
modes$mode # which regions are the modes?
hist[modes$mode, 1:4] # lower and upper bounds of the two modes
```

We highlight the two modes in the histogram, and the asterisks are the two true modes. The two modes in the Beta-trees histogram are close to the actual modes in the distribution. 

```{r, echo = F}
# function to plot the candidate modes
# INPUTS
# hist -- the histogram 
# modes -- the true location of the modes, should have two columns called "x" and "y"
# candidate_modes -- the candidate modes returned by our algorithm
PlotModes2D <- function(hist, modes, candidate_modes){
  colnames(hist) <- c("xmin","ymin","xmax","ymax",'density','lower','upper', "ndat",'level')
  hist <- as.data.frame(hist)
  g <- ggplot() +
        geom_rect(aes(xmin = xmin, xmax = xmax,ymin = ymin, ymax = ymax,fill = density), color = "black",data = hist)  +
    scale_fill_gradient2(low = "#f7fbff", mid = "#6baed6", high = "#08306b",aesthetics = "fill", midpoint = median(hist$density)) + 
        geom_rect(aes(xmin = xmin, xmax = xmax,ymin = ymin, ymax = ymax, fill = density), data=hist[candidate_modes, ],  color = "red") + 
    geom_point(aes(x = x, y= y), data=modes, shape = 8, color = "red", size = 2) +
        xlab("X") +  ylab("Y") +
        theme_bw() +  theme(text=element_text(size =18))
  
  return(g)
}
```

```{r, echo = F, fig.height = 3, fig.width= 4.5, fig.align='center'}
Mode2d <- data.frame(rbind(mu2d[[1]], mu2d[[2]])) 
colnames(Mode2d) <- c('x', 'y')
PlotModes2D(hist, Mode2d, modes$mode)
```


In the next section, we describe details of the `FindModes` function. 

## Identifying modes of a Beta-tree histogram

Let's rephrase the question of mode detection as the following: how can we tell that the two regions highlighted in red (let's call them $R_1$ and $R_2$) are two distinct modes? (If we know how to do this, then we can detect modes by iterating through all the regions , compare each one with the current list of distinct modes, and declare it a mode if it is distinct from all the current modes.) Suppose that $R_1$ and $R_2$ are distinct modes, then along every path that connects $R_1$ and $R_2$, there should exist at least one region $R$ whose average density is lower than those of both $R_1$ and $R_2$. Because the Beta-tree histogram returns a CI for the average density, we compare the lower and upper bound of the CI, that is, there should exist a region $R$ such that

$$
\mathrm{CI}_\mathrm{up}(R) < \min(\mathrm{CI}_\mathrm{low}(R_1), \mathrm{CI}_\mathrm{low}(R_2)),
$$
where $[\mathrm{CI}_\mathrm{low}(R), \mathrm{CI}_\mathrm{up}(R)]$ is the confidence interval of the average density of $R$. 

For example, we plot the confidence intervals for every region along the shortest path $R_1$ and $R_2$. As we can see, the upper confidence bound of the third  region is below the lower endpoints of both modes. Suppose this is true for every path, then we declare $R_1$ and $R_2$ are two distinct modes. 

```{r, echo = F}
adj <- compute_adjacency_mat(hist = hist, d = d) 
g <- igraph::graph_from_adjacency_matrix(adj, mode = "undirected")
```

```{r, echo = F}
# function to plot density along the shortest path between two regions A and B
# INPUTS
# hist -- the histogram 
# d -- data dimension
# g -- adjacency graph of the regions in histogram 
# A, B -- two regions
plotdensity <- function(hist, d, g, A, B){
  # find the shortest path between A and B
  path <- shortest_paths(g, from = A, to = B)$vpath[[1]]
  # make a bar plot of the density along the path
  density <- data.frame(
    id = factor(1:length(path)),
    density = hist[path, (2*d + 1)],
  lower = hist[path, (2*d + 2)],
  upper = hist[path, (2*d + 3)])
  # add two lines of the shorter of the lower CI of A and B 
  g <- ggplot(density) + 
    geom_point(aes(x = id, y = density)) + 
    geom_segment(aes(x = id, xend = id, y = lower, yend = upper)) + 
    geom_hline(yintercept = min(density$lower[1], density$lower[length(path)]),
               linetype = "dashed") + 
    ylab("Density") + 
    theme_bw() + scale_x_discrete(name ="Regions", 
                    labels =c("Mode 1", rep("", length(path)-2), "Mode 2"))
  
  return(g)
}
```

```{r, echo = F, fig.height = 3, fig.width= 4, fig.align='center'}
plotdensity(hist, d = d, g = g, A = modes$mode[1], B = modes$mode[2])
```

Note that we have only examined one path here, and to check every path, we use the function `is_connected()`. It returns `connected`  if the two regions are **not** distinct modes. The function has five input values: `i` and `j` are the indices of the two regions, `g` is a graph based on the adjacency matrix (we will describe later), `ci` are the lower and upper confidence bounds, and  `cutoff`  specifies the maximum length of path the function checks.

```{r}
BetaTrees::is_connected(i = modes$mode[1], j = modes$mode[2], g = g, ci = hist[,6:7], cutoff = 6)
```

We need the adjacency matrix in order to compute the path between two regions. The function `compute_adjacency_mat()` computes the adjacency matrix for all of the regions in the histogram. The two inputs of the function is a Beta-tree histogram `hist` and data dimension `d`. In the output adjacency matrix $A$, $A_{i,j} = 1$ if $i$-th and $j$-th regions are neighbors and 0 otherwise. We say that $R_1$ and $R_2$ are neighbors if 
$$
[x^{\mathrm{low}}_{1,j}, x^{\mathrm{up}}_{1,j}] \cap [x^{\mathrm{low}}_{2,j}, x^{\mathrm{up}}_{2,j}] \neq \emptyset,\quad \forall j=1,\ldots, d. 
$$
Here, we represent each region by the lower and upper bound in each coordinate, i.e., $R_1$ can be represented as $(x^{\mathrm{low}}_{1,1}, x^{\mathrm{up}}_{1,1})\times \ldots \times (x^{\mathrm{low}}_{1,d}, x^{\mathrm{up}}_{1,d})$. 

```{r}
adj <- compute_adjacency_mat(hist = hist, d = d) 
adj[1:5, 1:5] 
```

From the adjacency matrix, we compute an undirected graph `g` (using the function `graph_from_adjacency_matrix` from `igraph` package). 

```{r}
g <- igraph::graph_from_adjacency_matrix(adj, mode = "undirected")
```

To summarize, `FindModes` proceeds in the following steps: 

1. Order regions by the empirical density (in descending order). Initiate the list of modes by the region with the highest density. 

2. Iterate through every region by the order given in (1), test if it is a *distinct mode* from the current list of modes using `is_connected` function. If it is `unconnected` from all the current modes, then we add it to the list of modes.  

3. Return the list of modes. 




